\documentclass{article}

%\usepackage{colt11e}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{pseudocode}
\usepackage{dsfont}
\usepackage{graphicx}
\usepackage{url}

%\input{Macros}

\title{Java Program Execution Analysis Tool}

\author{
  Dongyang Zhang \\
  {\fontsize{10}{11}\selectfont University of Texas, Austin}\\ 
  \texttt{dyz@utexas.edu}
\and
  Jeremy Joachim \\
  {\fontsize{10}{11}\selectfont University of Texas, Austin}\\ 
  \texttt{jajoachim@gmail.com}
}

\begin{document}
\maketitle

\section{Introduction}

Software testing has always been a fickle yet vital task for product assurance.
As a result of growing software complexity, software testing has become a very expensive process for any programming language.
Java is one such language, and many profiling hooks to the Java Virtual Machine's (JVM) inner workings are made readily available for debugging tools to exploit.
Furthermore, the Java language has many constraints not found in complex langauges like C++ that allow code analysis to be more concise and straightforward.

\section{Problem Statement}

We aim to build a Java program execution profiling tool that instruments the Java bytecode using the ASM library.
The tool's main purpose will be to help view code coverage, especially in the context of test run, and map the code to a node graph.
Larger performance tests will also benefit from this tool by showing the hotspots in code.
This will allow the client to visualize what parts of the code should be optimized.

\section{Objectives}

Within the two main types of program analysis, static and dynamic, we wish to complete a small list of objectives.
In static analysis, we wish to display a basic block node map of the program.
This is to help the client visualize the branches in their project and the interactions in the object heirarchy.
This phase could benefit from a syntactic pathfinding tool to see if different nodes can connect, however the importance of such a function seems limited. 

Dynamic analysis is the main goal of this project.
The primary dynamic analysis objective in this project is to track method calls, method timings, and inter-method paths.
Additionally, the program will introduce a unique tool to display the difference between two executions of the same code.
Showing the difference between executions can show how the code reacts to different inputs.
If time persists, we would also like to perform CPU usage and memory footprint analysis on classes and methods within a program.
The following sections will describe our current plans for each implementation in both static and dynamic analyses.

\subsection{Static Analysis}

The profiler will parse all classes within a program and create a node map.
Each node will represent a basic block, but method calls will be in separate nodes.
Each node will be able to be expanded into bytecode instructions.
In the case of method call nodes, they will either jump to the method's node tree or simply display the bytecode that calls the method.

Pathfinding in a linked graph can be done with Dijkstra's algorithm.
The nodes could be either be weighted based on code complexity within each basic block, or all nodes have uniform weight.
We will only implement this functionality if time allows us.

\subsection{Dynamic Analysis}

Each method will keep a set of simple paths executed with each method call.
Knowing the amount of times each path was execution also shows how many times each node and edge is executed.
However, loops and repeated nodes will have to be handled in a special manner that has not yet been decided.
We are currently thinking of keeping them as a special node within a path that represets a subset nodes.

Method timing will be implemented with two 64 bit counters.  
One counter accumulates the total time spent in the method during execution (we will aim for nanosecond accuracy).
The other counter keeps track of how many calls to the method there are.
When the program needs to display the average time spent in a method, it only has to divide the two numbers.
Methods with recursive calls may yield skewed results.
Additionally, very short methods that will have runtimes in the timer's noise range will be excluded.
Their timings can be extrapolated from methods that call these smaller ones.

Finally, we will offer a display showing the difference between two executions of a program.
This will be implemented by first normalizing all of the timings of each method in both executions.
Next, a subtraction is performed between the two programs' method timings.
The difference in normalized timings will show the difference in program hotspots in two different executions.

CPU utilization and memory footprints are a bit more complicated to track; these low level statistics are not as readily available from the JVM.
There hooks that allow some limited functionality, but multithreaded applications will cause additional complexity.
We will only implement this functionality if time allows us.

\section{Related tools}

Java profiling tools have been an object of development for some time.
Jmap \cite{Oraclejmap}, developed by Oracle, focused on printing shared object memory maps or heap memory details.
(NEEDS IMRPOVEMENT) It is able to show object counts, increases of java class with little delay even for very large JVM heaps.
The same functionality can be found in other memory analzying tools such as in \cite{MAT}.

Another notable profiling tool is the VisualVM \cite{VisualVM} built in the JVM.
the creators described it as a visual tool integrating several command line JDK tools and lightweight profiling capabilities.
It enables gathering statistics throughout normal application execution rather than sampling information at intervals.
This is the kind of instrumentation our profiler will aim to do.
The downside that is the profiler is brute-force and it redefine most of the classes and methods.

BTrace \cite{BTrace} can let the user specify what type of information they want to gather by scripts.
The benefit is BTrace script is just a normal Java class containing some special annotations to specify where and how BTrace will instrument the application.
BTrace scripts are then compiled into standard class files by the compiler.
The drawback is the design makes it difficult to navigate the output format to process the data.
Later on, tools like EurekaJ \cite{EurekaJ} provide better visualization and parsing of the statistics.
(EITHER NEEDS ELABORATION OR OMISSION) Commercial tools like JProfiler \cite{JProfiler} support more extensive profiling including higher level profiling data, database profiling, analysis of memory leaks, QA capabilities as well as integrated thread profiling.

\section{Conclusion}
The purpose of our tool is to help programmers visualize the code coverage and code utilization of test cases.
In addition to these debugging features, the tool will also provide code hot spots during each test case.
This can help programming find areas of code that would be best to optimize.
(A BIT MORE HERE)

\bibliographystyle{plain}
\bibliography{proposal}
\end{document}
